Temporal ordering and registration of images in studies of developmental dynamics

### 2015-01-28 image_analysis_paper_development2/manuscript.tex
updated all files

@@

Imaging studies provide unique insights into the dynamics of pattern formation and morphogenesis in developing tissues.
%
Developmental progress is commonly reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific developmental system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in  \textit{Drosophila} embryos.

@@

We present a robust algorithmic approach to registration and temporal ordering.
%
In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
%
Our approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
%
Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze two data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our first data set, from a live imaging study where the correct rotational orientation and temporal order are independently known, will be used to validate our approach.
%
Our second data set consists of images from fixed embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.

@@

\begin{figure}[t]
\includegraphics{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. (A) Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. (B) The vectors from A, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in A: the differences between a pair of angles in B is the same as the pairwise angle in A. (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \SI), so that close data points are connected by darker (``stronger'') edges. (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
%\label{fig:schematics}
\customlabel{fig:schematics}{2}
\customlabel{subfig:synch1}{\ref{fig:schematics}A}
\customlabel{subfig:synch2}{\ref{fig:schematics}B}
\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
\end{figure}

Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{subfig:synch1}, where each image is represented as a vector, and the goal is to align the set of vectors.
%
We first compute the angles needed to align pairs of vectors (or images).  
%
In general, this requires no template function \citep{ahuja2007template} or image landmarks \citep{ian1998statistical}.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{subfig:synch2}.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can successfully register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%
Such a manifold can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{subfig:dmaps1}:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

\begin{figure*}[t]
\raisebox{0.75cm}{\textsf{A}}
\hfill
\includegraphics[width=17.5cm]{drosophila_live_imaging_scrambled}

\raisebox{0.75cm}{\textsf{B}}
\hfill
\includegraphics[width=17.5cm]{drosophila_live_imaging_ordered}

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.75cm}{
{ \textsf{ 
\begin{tabular}{ |p{1.25cm} | p{1.5cm}|  p{1.75cm}| }
\hline
Movie Index & Average Angle Error & Rank Correlation Coefficient \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\ 
2 & 4.98$^{\circ}$ & 0.9994 \\ 
3 & 6.47$^{\circ}$ & 0.9795 \\ 
4 & 7.03$^{\circ}$ & 0.9953 \\ 
5 & 3.04$^{\circ}$ & 0.9959 \\ 
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies. }
\end{figure*}

@@

\fig~\ref{subfig:movie_unregistered_unordered} shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo (see \fig~\ref{fig:full_movie_data}).
%
Each image shows an optical cross-section of a vertically oriented developing embryo, with the nuclei labeled by Histone-RFP.
%
Each frame was arbitrarily rotated, and the order of the frames was scrambled.
%
The task is now to register these images and order them in time to reconstruct the developmental trajectory.

We used vector diffusion maps to register and order the images. 
%
\fig~\ref{subfig:movie_registered_ordered} shows the images from \fig~\ref{subfig:movie_unregistered_unordered}, now registered and ordered; the real time for each frame is also indicated.
%
With a small number of exceptions, the recovered ordering is consistent with the real time dynamics. 
%
\fig~\ref{subfig:movie_angle_corr} and \fig~\ref{subfig:movie_rank_corr}  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set: 
%
Both the angles and the ranks are recovered with a high degree of accuracy.

To assess the robustness of our proposed methodology, we repeated this procedure with four additional data sets (see \fig~\ref{fig:all_movies}) extracted from independent live imaging studies spanning the same developmental time period. 
%
The results are shown in \fig~\ref{subfig:movie_error_table}. 
%
The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.

In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{subfig:dmaps1}) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images. 
%% 
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
%
In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

\begin{figure*}[t]
\begin{minipage}[c]{13.75cm}
\vspace{0cm}
\raisebox{0.5cm}{\textsf{A}}
\includegraphics[width=13.5cm]{zebrafish_scrambled}

\raisebox{0.5cm}{\textsf{B}}
\includegraphics[width=13.5cm]{zebrafish_ordered}
\end{minipage}
%
\hfill
%
\begin{minipage}[c]{3.5cm}

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.15cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. Registered and ordered zebrafish data.}
\end{figure*}

@@

As in the live imaging {\em Drosophila} example, the frames were randomly rotated and shuffled. 
%
We then used vector diffusion maps to register and order the frames. 
%
The results are shown in \fig~\ref{fig:zebrafish}. 
%
The recovered rotations and order are consistent with the expected developmental dynamics. 
%
Quantitatively, the rank correlation coefficient for this data set is $0.9935$, and the average error in the recovered angle is $4.23^\circ$.

@@

\raisebox{0.5cm}{\textsf{C}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
\end{minipage}
%
\hfill
%
\raisebox{3.25cm}{\textsf{D}}
\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
%
\caption{Analyzing images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.97$. }
\end{figure*}

@@

Finally, we applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
%
\fig~\ref{subfig:fixed_images_unregistered_unordered} shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
%
This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provides information about tissue morphology and the spatial distribution of two regulatory proteins.
%
Each image shows an optical cross-section of a {\em different} vertically oriented embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
%
The nuclei (gray) were labeled with DAPI, a DNA stain.
%
Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \citep{Lim2013kinetics}.

\fig~\ref{subfig:fixed_images_registered_ordered} shows the selected images in \fig~\ref{subfig:fixed_images_unregistered_unordered}, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set (see \ref{fig:full_fixed_data}).
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI). 
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:fixed_images_average_trajectory}, serve as a summary of the stereotypic developmental dynamics.
%
We can now easily see the developmental progression, which is consistent with the known dynamics: 
dpERK first appears as two lateral peaks at the ventral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
%
During mesoderm invagination, the two ventral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
%
The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
%
At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental features (e.g., correlation of age with body size).

To evaluate the quality of our registration and ordering, we can use prior knowledge about the developmental system. 
%
The Twi signal is known to form a single peak at the ventralmost point of the embryo. 
%
We found that the standard deviation in the location of this peak in the set of registered images was $\sim$8$^\circ$,
indicating that the algorithm successfully aligns the ventralmost points of the images. 
%
Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our temporal ordering. 
%
We compared the ordering obtained from vector diffusion maps to the ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.  
%
The ranks from the ordering provided by the embryologist, which we will refer to as the ``expert rank'', are indicated for the images in \fig~\ref{subfig:fixed_images_registered_ordered} , and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.

Manually ordering the images is nontrivial for researchers who are unfamiliar with the developmental progression, and can be tedious and time-consuming for those who are.
%
In contrast, our method requires relatively little computational time; registering and ordering this fixed imaging data set took $26$~seconds on an Intel Core i7 2.93 GHz processor. %for $100$ images of $100 \times 100$ pixels, using 10$^{\circ}$ rotational discretizations.
%
The computational time is a function of the number of images in the data set, the number of pixels in each point, and the angular resolution to compute the pairwise rotations (see \SI). 
%
Empirically, the CPU time required was $\mathcal{O}(n^{1.5})$ in the number of images, and $\mathcal{O}(n)$ in the number of pixels and the angular resolution.
%
Furthermore, the computation of the pairwise rotational alignments, which is the most time-intensive portion of the calculation, is trivially parallelizable.
%
The requisite user intervention and parameter tuning required for our method is relatively minimal: Images must first be preprocessed so that the Euclidean distance between the pixels is informative.
%
Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined.
%
We found that the results are robust to both of these parameters, and that consistent parameter values served well for all data sets presented here. 
%
%In our examples, we use 10$^{\circ}$ angle discretizations for the pairwise alignments, and one-tenth of the median pairwise distance for the kernel scale. 
% (5$^{\circ}$--30$^{\circ}$ angle discretizations and kernel scales of one-fifth to one-twentieth of the median pairwise distances all yielded acceptable results)
%
Overall, the tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.

@@

\begin{figure*}[t]
\begin{minipage}[c]{9cm}
\vspace{0cm}

\raisebox{6.75cm}{\textsf{A}}
\includegraphics[width=8cm]{wing_disc_example}
\end{minipage}
%
\begin{minipage}[c]{8cm}
\caption{Registered and ordered three-dimensional wing-disc data}
\end{minipage}

\raisebox{0.7cm}{\textsf{B}}
\includegraphics[width=\textwidth]{wing_disc_ordered}

\end{figure*}

>>>>

\usepackage{epstopdf}

@@

Imaging studies provide unique insights into the dynamics of pattern formation and morphogenesis in developing tissues.
%
Developmental progress is commonly reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific developmental system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in three different model systems in developmental biology.

@@

We present a robust algorithmic approach to registration and temporal ordering.
%
In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
%
Our approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
%
Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze four data sets common in developmental biology. 
%
Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis; here, the correct rotational orientation and temporal order are independently known and these data sets will be used to validate our approach.
%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent. 
%
Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods can extend to three-dimensional imaging.

@@

\begin{figure}[t]
\includegraphics{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. (A) Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. (B) The vectors from A, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in A: the differences between a pair of angles in B is the same as the pairwise angle in A. (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \SI), so that close data points are connected by darker (``stronger'') edges. (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
\label{fig:schematics}
%\customlabel{fig:schematics}{2}
%\customlabel{subfig:synch1}{\ref{fig:schematics}A}
%\customlabel{subfig:synch2}{\ref{fig:schematics}B}
%\customlabel{subfig:dmaps1}{\ref{fig:schematics}C}
%\customlabel{subfig:dmaps2}{\ref{fig:schematics}D}
\end{figure}

Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the set of vectors.
%
We first compute the angles needed to align pairs of vectors (or images).  
%
In general, this requires no template function \citep{ahuja2007template} or image landmarks \citep{ian1998statistical}.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{fig:schematics}B.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can successfully register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%
Such a manifold can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

\begin{figure*}[t]
\raisebox{0.75cm}{\textsf{A}}
\includegraphics[width=17cm]{drosophila_live_imaging_scrambled}

\raisebox{0.75cm}{\textsf{B}}
\includegraphics[width=17cm]{drosophila_live_imaging_ordered}

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.75cm}{
{ \textsf{ 
\begin{tabular}{ |p{1.25cm} | p{1.5cm}|  p{1.75cm}| }
\hline
Movie Index & Average Angle Error & Rank Correlation Coefficient \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\ 
2 & 4.98$^{\circ}$ & 0.9994 \\ 
3 & 6.47$^{\circ}$ & 0.9795 \\ 
4 & 7.03$^{\circ}$ & 0.9953 \\ 
5 & 3.04$^{\circ}$ & 0.9959 \\ 
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies. }
\label{fig:drosophila_live_imaging}
\end{figure*}

@@

\fig~\ref{fig:drosophila_live_imaging}A shows selected images from this live imaging data set, which contains $40$ consecutive frames taken at $30$~second time intervals at a fixed position within a single embryo (see \fig~\ref{fig:full_movie_data}).
%
Each image shows an optical cross-section of a vertically oriented developing embryo, with the nuclei labeled by Histone-RFP.
%
Each frame was arbitrarily rotated, and the order of the frames was scrambled.
%
The task is now to register these images and order them in time to reconstruct the developmental trajectory.

We used vector diffusion maps to register and order the images. 
%
\fig~\ref{fig:drosophila_live_imaging}B shows the images from \fig~\ref{fig:drosophila_live_imaging}A, now registered and ordered; the real time for each frame is also indicated.
%
With a small number of exceptions, the recovered ordering is consistent with the real time dynamics. 
%
\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set: 
%
Both the angles and the ranks are recovered with a high degree of accuracy.

To assess the robustness of our proposed methodology, we repeated this procedure with four additional data sets (see \fig~\ref{fig:all_movies}) extracted from independent live imaging studies spanning the same developmental time period. 
%
The results are shown in \fig~\ref{fig:drosophila_live_imaging}E. 
%
The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.

In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images. 
%% 
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignments and the errors in the recovered angles were consistently less than $10^{\circ}$.
%
In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

\begin{figure*}[t]
\begin{minipage}[c]{13.75cm}
\vspace{0cm}
\raisebox{0.5cm}{\textsf{A}}
\includegraphics[width=13cm]{zebrafish_scrambled}

\raisebox{0.5cm}{\textsf{B}}
\includegraphics[width=13cm]{zebrafish_ordered}
\end{minipage}
%
\hfill
%
\begin{minipage}[c]{3.75cm}

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.25cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a time-lapse movie of zebrafish embryogenesis. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9936$.}
\label{fig:zebrafish}
\end{figure*}

@@

As in the live imaging {\em Drosophila} example, the frames were randomly rotated and shuffled. 
%
We then used vector diffusion maps to register and order the frames. 
%
The results are shown in \fig~\ref{fig:zebrafish}. 
%
The recovered rotations and order are consistent with the expected developmental dynamics. 
%
Quantitatively, the rank correlation coefficient for this data set is $0.9936$, and the average error in the recovered angle is $4.23^\circ$.

@@

\raisebox{0.5cm}{\textsf{C}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
\end{minipage}
%
\hfill
%
\raisebox{3.25cm}{\textsf{D}}
\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
%
\caption{Analyzing images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.9716$. }
\label{fig:drosophila_fixed_images}
\end{figure*}

@@

Finally, we applied our approach to a data set where the true rotational orientation and temporal order was not known {\it a priori}.
%
\fig~\ref{fig:drosophila_fixed_images}A shows selected images from a set of $120$ images (see \fig~\ref{fig:full_fixed_data}) which cover a thirty minute time interval spanning late cellularization through gastrulation.
%
This data set is more complex than the live imaging data sets in that it contains significantly more images, each of which provides information about tissue morphology and the spatial distribution of two regulatory proteins.
%
Each image shows an optical cross-section of a {\em different} vertically oriented embryo at a different rotational orientation and fixed at a different (and unknown) developmental time.
%
The nuclei (gray) were labeled with DAPI, a DNA stain.
%
Embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor which specifies the cells of the future muscle tissue.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, shown in red), an enzyme that, in this context, specifies a subset of neuronal cells \citep{Lim2013kinetics}.

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set (see \ref{fig:full_fixed_data}).
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI). 
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.
%
We can now easily see the developmental progression, which is consistent with the known dynamics: 
dpERK first appears as two lateral peaks at the ventral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
%
During mesoderm invagination, the two ventral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
%
The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
%
At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{fig:drosophila_fixed_images}C.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental features (e.g., correlation of age with body size).

To evaluate the quality of our registration and ordering, we can use prior knowledge about the developmental system. 
%
The Twi signal is known to form a single peak at the ventralmost point of the embryo. 
%
We found that the standard deviation in the location of this peak in the set of registered images was $\sim$8$^\circ$,
indicating that the algorithm successfully aligns the ventralmost points of the images. 
%
Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our temporal ordering. 
%
We compared the ordering obtained from vector diffusion maps to the ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.  
%
The ranks from the ordering provided by the embryologist, which we will refer to as the ``expert rank'', are indicated for the images in \fig~\ref{fig:drosophila_fixed_images}B , and the rank correlation (see \fig~\ref{fig:drosophila_fixed_images}D) shows that our ordering is consistent with the expert ordering.

@@

\begin{figure*}[t]
\raisebox{5.5cm}{\textsf{A}}
\includegraphics[width=6cm]{wing_disc_example}
%
\hfill
%
\raisebox{5.5cm}{\textsf{B}}
\includegraphics[width=10cm]{wing_disc_ordered}
%
\caption{Three-dimensional wing disc data. (A) Maximum projections of an example three-dimensional image. Projections along the $x$-, $y$-, and $z$-axes are shown. (B) Example three-dimensional images, ordered using diffusion maps. The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427.}
\label{fig:wing_disc}
\end{figure*}

Up to this point, we have focused on two-dimensional imaging studies. 
%
However, it is not uncommon to encounter three-dimensional imaging studies. 
%
To illustrate how our methods can be extended to three-dimensional data, we analyzed a wing disc data set.
